---
title: "Learning Checks Lesson 22"
---

```{r include=FALSE}
source("../_startup.R")
```

## 22.1 (**Obj 20.3**)

Consider these three data frames:

```{r}
One <- sample(dag01, size=25)
Two <- do(10) * {
  lm(y ~ x, data = sample(dag01, size=25)) %>%
    coefficients()
  }
Three <- Two %>% 
  summarize(mx = mean(x), sx = sd(x))
```

1. Both `One` and `Two` have columns called `x`, but they stand for different things. Explain what the unit of observation is and what the values in `x` represent..
2. `Three` does not have a column named `x`, but it is a summary of the `x` column from `Two`. What kind of summary.

::: {.callout-tip}
## Solution

1. In `One`, the `x` column contains the simulated of the `x` variable from `dag01`. The unit of observation is a single case, for instance a person for whom observations were made of `x` and `y`. The simulation involves generating 25 rows of data: one row for each of 25 people.
2. In `Two`, the `x` column is the regression coefficient on `x` from the simulation. Each row of `Two` corresponds to one trial in which regression is being performed on a sample of size 25 of simulated data from `dag01`. 
3. `Three` is a summary of the 10 trials in `Two`. The columns, named `mx` and `sx`, tell about the distribution of `x` across all the trials. 
:::


--------
