---
title: "Learning Checks Lesson 21"
---

```{r include=FALSE}
source("../_startup.R")
```

## 21.1

The following command will generate a data frame with 1000 rows from `dag00` and calculate the variance of the `x` and `y` variables:

```{r}
sample(dag00, size=1000) %>%
  summarize(vx = var(x), vy = var(y))
```

Compare this result to the DAG tilde expressions

```{r}
dag00
```

In the tilde expressions, `exo(2)` means to generate noise of magnitude 2.0.

1. Is the argument to `exo()` specified in terms of the variance or the standard deviation?
2. The tilde expression for `x` specifies that the constant `5` is to be added to `exo(2)`. Similarly, the constant `-7` is added to `y`.  How do these constants relate to the calculated magnitudes of `x` and `y`?

### Solution

1. The standard deviation. For instance, `x` has noise of magnitude 2. The variance of `x` is 4, the square of 2.
2. The standard deviation (and therefore the variance) ignore such added constants.

--------



## 21.2

@sec-signal-and-noise introduces the idea that variables consist of components. A simple breakdown is into two components: i. the part of the variable that is determined by other variables in the system ("signal") and ii. the random part of the variable ("noise"). The section uses `dag01` as an illustration of how a variable can be partly determined and partly random noise.

i. Write and execute a command that will generate 500 rows of simulated data from `dag01` and will calculate the standard deviation of `x` and of `y`. 

ii. What's the magnitude of `x` in the simulated data? What's the magnitude of `y`?

iii. Does this change if you use data with 1000 or 20000 rows?




### Solution

i. `sample(dag01, size=500) %>% summarize(sx = sd(x), sy=sd(y))`
ii. The standard deviation of `x` is about 1, the standard deviation of `y` is about 1.8.
iii. No, the values are roughly the same regardless of the size of the sample.

----------



## 21.3

[DRAW several DAG-like graphs, one of which should be undirected in all edges, one should be undirected on one or two edges (but not all), and one should be cyclic and another acyclic.]

Referring to the graphs in the figure, say which ones are DAGs. If a graph is not a DAG, say whether that's because it's not **directed** or because it's not **cyclic**.

### Solution



## 21.A

A DAG (directed acyclic graph) is a mathematical object used to state hypothetical causal relationships between variables. Explain briefly (e.g. a few sentences overall) what each of the words "directed," "acyclic," and "graph" mean in the context of a DAG.

::: {.callout-note}
## Solution

1. A graph is a relationship between discrete elements (called "nodes" abstraction) each of which for us represents a hypothetical quantity, that is, a variable. In addition to the nodes, the graph contains edges which represent the connections between variables.

2. A directed graph is one whose arrows have a direction. For instance $A \rightarrow B \leftarrow C$ means that $A$ and $C$ together cause $B$, but $B$ has no influence on $A$ and $C$.

3. An acyclic graph is one where it is impossible to start on any given node and, by following the directed edges, return back to that node.
:::

----------

## 21.B

Draw these DAGs:

1. "April showers bring May flowers."

2. "Price of a rug is determined by the size and the quality of materials."

3. "The weight of an automobile is reflected in the MPG fuel economy, as is the speed of the car, and inflation level of the tires."

4. "Plants tend to grow in the direction of the sunlight."

5. "An ice-cream shop owner needs to plan staffing based on the season, day of the week, and holidays."

----------

## 21.C




----------

## 21.4

Generate simulated data from `dag01` with 1000 rows. Fit the regression model `y ~ x` to the data and examine the coefficients. 

1. How do the coefficients relate to the tilde expressions that define `dag01`?

2. Instead of using the regression model `y ~ x`, where `y` is the response variable, try the regression model `x ~ y`. Do the coefficients from `x ~ y` correspond in any simple way to the tilde expressions that define `dag01`?

### Solution

```{r}
sample(dag01, size=1000) %>%
  lm(y ~ x, data = .)
```

The intercept corresponds to the additive constant (4) in the `y` tilde expression. The `x` coefficient corresponds to the multiplier on `x` in the tilde expression. 

The formula for `x` isn't reflected by the coefficients.


Using `x` as the response variable:

```{r}
sample(dag01, size=10000) %>%
  lm(x ~ y, data = .)
```

These coefficients do not appear in the `dag01` tilde expressions.

--------

## 21.5 (Objective 21.2)

You are trying to understand why automobile fuel economy varies from model to model. Using the `mtcars` data frame (documentation at `help("mtcars")`) ...

i. What's an appropriate choice of a response variable?
ii. Pick two explanatory variables of interest to you. Build an appropriate model from the data, extracting the `coefficients()` of the model. Explain what the coefficients mean in everyday terms that your cylinder-head uncle would approve of.
iii. Which of the other variables are covariates? 
iv.Pick a covariate that your intuition suggests would be important. Include that covariate in the model from (ii) and say whether the covariate shows up as important in the model coefficients.

::: {.callout-note}
## Solution

iii. All of the other variables are covariates. A covariate is merely a potential explanatory variable that you are not directly interested in. Of course, not all covariates play an important role in the system.
:::

----------

## 21.6 (Obj 21.3)

Generate a sample of size $n=100$ from `dag03`. Use the data to construct a model of `y` versus `x`. But instead of using `coefficients()` to look at the model coefficients, use `confint()`. While `coefficients()` reports a single value for each coefficient, `confint()` reports a plausible interval for coefficients that is consistent with the data. 

i. For $n=100$, how wide is the interval reported by `confint()` on the `x` coefficient.
ii. Repeat the process of sampling and modeling, but this time use $n=400$. How wide is the interval reported by `confint()` on the `x` coefficient.
iii. Again repeat the process of sampling and modeling, this time using $n=1600$. iv. Does the interval reported depend systematically on the size $n$ of the sample? Describe what pattern you see.

--------

## 21.7

A short report from the British Broadcasting Company (BBC) was headlined "[Millennials' pay 'scarred' by the 2008 banking crisis](https://www.bbc.com/news/business-47106711)."

> Pay for workers in their 30s is still 7% below the level at which it peaked before the 2008 banking crisis, research has suggested. The Resolution Foundation think tank said people who were in their 20s at the height of the recession a decade ago were worst hit by the pay squeeze. It suggested the crisis had a lasting "scarring" effect on their earnings.

> The foundation said people in their 30s who wanted to earn more should move to a different employer. The research found those who stayed in the same job in 2018 had real wage growth of 0.5%, whereas those who found a different employer saw an average increase of 4.5%.

The phrase "should move" in the second paragraph of the report suggests causation. A corresponding DAG would look like this:

- move_to_new_employer $\rightarrow$ higher_wage

For the sake of discussion, let's add two more variables to the DAG:

- effectiveness---standing for how productive the employee is.
- qualifications---standing for whether the employee is a good candidate attractive to potential new employers.

Construct a DAG with all four variables that represent plausible causal connections between them. The DAG should not contain a direct causal connection  between move_to_new_employer and higher_wage.


If the world worked this way, would it necessarily be good advice to switch to a new employer with the aim of earning a higher wage?


