---
title: "Learning Checks Lesson 29"
---

```{r include=FALSE}
source("../_startup.R")
```

## 29.A {#sec-LC-29-A}



The `math300::Hill_racing" data frame records 2236 winning times (in seconds) in Scottish hill racing competitions. Consider this model of the winning time as a function of the race distance (km) and the total climb (meters):

```{r}
mod <- lm(time ~ distance + climb, data=Hill_racing)
```

The `model_eval()` function provides a convenient way to evaluate the model output (`.output`) for each of the rows in a data frame and, at the same time, calculates row-by-row residuals (`.resid`) and prediction errors (`.lwr` and `.upr`). *Make sure to take not of the starting periods on the names.*

```{r}
model_eval(mod) %>% head()
```




The RMS residual from the model can be calculated this way:

```{r}
model_eval(mod) %>%
  summarize(rms = sqrt(mean(.resid^2)))
```
1. What are the units of the RMS residual?

2. Modify the calculation to compute the sum-of-square residuals. Report the result numerically. Be sure to say what are the units.

3. What are the units of the effect size on time with respect to climb? 

4. What are the units of the effect size on time with respect to climb? 

::: {.callout-note}
## Solution

1. RMS residual has the same units as the response variable. In this case, that's the time to run the race, with units "seconds."

2. SS residual has units that are the square of the respond variable, in this case "square-seconds." 

3. Recall that the effect size on the response with respect to an explanatory variable has the units of the response variable divided by the units of the explanatory variable. The `climb` variable has units of meters, so the effect size has units "seconds/meters."

4. seconds/km

:::

---------

## LC 29.B


Which of the following models are **not** nested within `time ~ distance + climb`?

a. `time ~ 1`
b. `time ~ distance + sex`
c. `time ~ distance`
d. `time ~ climb`

::: {.callout-note}
## Solution

The model `time ~ distance + sex` is not nested in `time ~ distance + climb`. 

Note that `time ~ 1` is indeed nested in `time ~ distance + climb`. The `1` corresponds to the intercept.

:::

--------

## 29.C 

In LC -@sec-LC-29-A you calculated the RMS residuals and the sum-of-square residuals by wrangling the results from `mod_eval()`. That's a perfectly good way to do things, but the work becomes tedious when there are multiple models you want to compare.

For convenience, there is a `compare_model_residuals()` command, which can calculate the RMS residual or sum-of-square residual for each of a set of models. All the models must have the same response variable.

```{r}
Hill_racing %>% 
  compare_model_residuals(time ~ 1, 
                          time ~ distance + climb, 
                          time ~ distance + climb + sex,
                          time ~ distance,
                          measure = "RMS"
                          )
```

It happens that all of the models in the command are a nested set. Re-order the models so that each model nests inside the following model, that is, from smaller model to bigger model. 

1. Do the RMS residuals for the nested models increase or decrease when moving from a smaller model to a larger model?

2. You can calculate the sum-of-square residual by using the argument `measure="SS"`. Do the sum-of-square residuals for the nested models increas or decrease when moving from a smaller model to a larger model.

3. You can calculate R^2^ by using the argument `measure="R2"`. Do the R^2^ for the nested models increase or decrease when moving from a smaller model to a larger model.

::: {.callout-note}
## Solution

1. Decrease
2. Decrease
3. Increase. R^2^ tells you how big the model output is compared to the response variable. 1-R^2^ tells you how big the residuals are compared to the response variable.
:::

--------

## LC 29.D

::: {.callout-warning}
## Still a draft

Look at `dag07`. Notice that `d` is not connected to any of the other variables.

Generate a sample of size $n=6$. Compare the sum of square residual (in sample) from the nested  models `c ~ 1`, `c ~ a` `c ~ a + b`, and `c ~ a + b + d`. (Use the `compare_model_residuals()` using the argument `method="SS"`.

Which, if any, of the variables `a`, `b`, or `d` reduces the in-sample sum-of-squared residuals compared to the previous model.


::: {.callout-note}
## Solution

```{r}
compare_model_residuals(dag07, c ~ 1, c ~ a, c ~ a + b, c ~ a + b + d, 
                        n=6, measure="R2")
```
:::

Out of sample, the useless covariate often *increases* the SS error.

--------

## LC 29.1

In `dag04`, build models to predict `c` from the other variables. Does one of those variables "block" the others? 

- Explain how you know this from your models. Try to give an answer in everyday language as well.
- Repeat but use a very small sample size, say $n=5$. Has your conclusion about blocking changed? Explain why.

::: {.callout-note}
## Solution
```{r}
compare_model_residuals(dag04, c~ 1, c ~ d, c~ b + d, c ~ a + b + d, n=50)
```

`d` seems to block effect of `a` and `b` on `c`.


```{r}
compare_model_residuals(dag04, c~ 1, c ~ d, c~ b + d, c ~ a + b + d, n=5)
```
:::

## LC 29.2

We are using in-sample testing because that is often the case in the model-building stage. However, in the model-**using** stage, things are different. You will be making predictions of new cases, that is, out-of-sample.

For out-of-sample, when working with new data, it's not just a matter of being tricked into thinking covariates are useful when they're not. Using irrelevant covariates can be genuinely harmful to the predictions.

Compare these in-sample and out-of-sample results. 

```{r}
set.seed(101)
compare_model_residuals(dag07, d ~ 1, d ~ c, d~ b + c, d ~ a + b + c, n=4)
set.seed(101)
compare_model_residuals(dag07, d ~ 1, d ~ c, d~ b + c, d ~ a + b + c, n=4, 
                        testing = "out-of-sample")
```

What do you see in the results that tells you that incorporating irrelevant covariates hurts the out-of-sample predictions?




--------

## LC 29.3

::: {.callout-warning}
## In draft

`openintro::teacher`. What's the `base` pay difference between a teacher with an MA and a BA degree? What's a confidence interval on this effect size? How does the confidence interval change if you include `years` as a covariate.
:::

--------

## LC 29.4

::: {.callout-warning}
## In draft

`openintro::census` Predict log personal income based on other variables. Eat variance using the `total_family_income` variable.

```{r}
mod <- lm(log10(total_personal_income) ~ log10(age) + sex + marital_status + log10(total_family_income), data = openintro::census %>% filter(total_personal_income > 0, total_family_income > 0))
anova(mod)
gf_jitter(total_personal_income ~ total_family_income | sex, 
         data =openintro::census %>% filter(total_personal_income > 3000),
         color=~marital_status, alpha=0.3) %>% 
  gf_refine(scale_y_log10(), scale_x_log10())
```


:::

--------

## 29.5

::: {.callout-warning}
## Still in draft

`openintro::starbucks`  where do the calories come from?  Find effect size of, say, protein on calories. Then see what happens if you use carbohydrates as a covariate.
```

```{r}
lm( calories ~ protein, data = openintro::starbucks) %>% confint()
lm( calories ~ fat + carb + fiber + protein, data = openintro::starbucks) %>% confint()
```

::: {.comment-warning}
## In draft

Maybe come back to this in confounding lesson. Look for components that tend to go together

```{r}
with(openintro::starbucks, cor(fat, protein))
with(openintro::starbucks, cor(fiber, protein))
lm( calories ~ fiber , data = openintro::starbucks) %>% confint()
lm( calories ~ protein, data = openintro::starbucks) %>% confint()
lm( calories ~ protein + fiber, data = openintro::starbucks) %>% confint()

```
:::

--------




