## Math 300Z: In-class group activity

```{r include=FALSE}
library(math300)
```

## Adjusting, visually

These are two graphs of the data from `Clock_auction` showing the relationship between the winning price and the number of bidders. (I've simplified the number of bidders to two categories.) The age of the clock is a covariate. The large dots show the mean age and mean price of the clocks in those auctions with 10 or more bidders versus 9 or fewer bidders.

```{r}
#| code-fold: true
Clock_auction <- Clock_auction |> mutate(nbidders = ifelse(bidders >= 10, "10 or more", "9 or fewer"))
Stats <- Clock_auction |> group_by(nbidders) |>
  summarize(mp = mean(price), mage = mean(age))
```

```{r warning=FALSE}
mod1 <- lm(price ~ nbidders, data=Clock_auction) 

model_plot(mod1, x=age, color=nbidders) |>
  gf_point(mp ~ mage, color=~nbidders, data=Stats, size=3)
```

**Part A**. In the model without `age` as a covariate, what is the difference in mean prices for the 10-or-more-bidders group versus the 9-or-fewer-bidders group?


**Part B**. Now the picture when including `age` as a covariate. Adjusting for `age`, what is the difference in mean prices for the 10-or-more-bidders group versus the 9-or-fewer-bidders group?

```{r warning=FALSE}
mod2 <- lm(price ~ nbidders + age, data=Clock_auction) 

model_plot(mod2, x=age, color=nbidders) |>
  gf_point(mp ~ mage, color=~nbidders, data=Stats, size=3)
```
 
**Part C**. Here are confidence intervals for the two models graphed above. Explain what about these coefficients matches the conclusions you got in Parts (A) and (B)? 
  
```{r}
mod1 |> conf_interval()
mod2 |> conf_interval()
```

-----

*This activity was inspired by schematic diagrams in Milo Schield's *Statistical Literacy: Seeing the story behind the statistics*, 2011, pp. 224-5.*
