---
title: "List of learning checks"
---

This list is assembled from the individual-lesson learning check files in the LC/ directory. Make any changes in that directory.

::: {.callout-note}
## In draft: Data, temporarily

Several of the problems refer to datasets in the `LC/data` subdirectory and documented in `LC/data/explain.qmd`. These should eventually be moved to the `math300` package. [Actually, not so many. Several of the examples from *CSP* refer to csv files, but those data are also available through the `openintro` or other packages.]
:::

## Non-technical resources

These are resources that can be used for an essay/reading exercise or class discussion. 

### `pudding.cool`

Examples:

- [Chinese censorship of the *Big Bang Theory*](https://pudding.cool/2022/08/censorship/) 
- [Comparing pocket sizes on women's and men's clothing](https://pudding.cool/2018/08/pockets/)
- [How artists get paid from music streaming](https://pudding.cool/2022/06/streaming/)
- [Randomness and age](https://pudding.cool/2022/04/random/). Perhaps suitable for Lesson 38 on "false discovery."
- [Coughing and the Academy awards](https://journal-doi.org/10.731/pcbi.1007742/) (The link was through <pudding.cool>.)
- [NBA draft](https://pudding.cool/2022/02/redraft/)
- [Writing plainly](https://pudding.cool/2022/02/plain/)
- many others!

### NYT "What's going on with this graph?"

[Link to index](https://www.nytimes.com/column/whats-going-on-in-this-graph)

- [Foreign born people as a fraction of the US Population](https://www.nytimes.com/2022/10/06/learning/whats-going-on-in-this-graph-oct-19-2022.html)
- [Tree rings and droughts](https://www.nytimes.com/2022/09/15/learning/whats-going-on-in-this-graph-sept-21-2022.html)
- many others!

### Exercises from *CPS*

- Section 29.6 Exercises 
    1. Linear and nonlinear correlations
    2. R^2^


## Lesson 19

```{r child="LC/LC-lesson19.qmd"}
```

## Lesson 20


```{r child="LC/LC-lesson20.qmd"}
```

## Lesson 21


```{r child="LC/LC-lesson21.qmd"}
```

## Lesson 22


```{r child="LC/LC-lesson22.qmd"}
```

## Lesson 23


```{r child="LC/LC-lesson23.qmd"}
```

## Lesson 24


```{r child="LC/LC-lesson24.qmd"}
```

## Lesson 25


```{r child="LC/LC-lesson25.qmd"}
```

## Lesson 26


```{r child="LC/LC-lesson26.qmd"}
```

## Lesson 27

```{r child="LC/LC-lesson27.qmd"}
```

## Lesson 28


```{r child="LC/LC-lesson28.qmd"}
```

## Lesson 29


```{r child="LC/LC-lesson29.qmd"}
```

## Lesson 30


```{r child="LC/LC-lesson30.qmd"}
```

## Lesson 31


```{r child="LC/LC-lesson31.qmd"}
```

## Lesson 32


```{r child="LC/LC-lesson32.qmd"}
```

## Lesson 33


```{r child="LC/LC-lesson33.qmd"}
```

## Lesson 34


```{r child="LC/LC-lesson34.qmd"}
```

## Lesson 35


```{r child="LC/LC-lesson35.qmd"}
```

## Lesson 36


```{r child="LC/LC-lesson36.qmd"}
```

## Lesson 37


```{r child="LC/LC-lesson37.qmd"}
```

## Lesson 38


```{r child="LC/LC-lesson38.qmd"}
```

