---
title: 'Lesson 36: Worksheet'
author: "Jane Doe"
---

```{r include=FALSE}
library(math300)
```

We have been constructing and using **confidence intervals** since Lesson 23. Recall that each **coefficient** of a model has its own confidence interval.

On many occasions, we have been interested in whether a connection is detectable between the response and a particular explanatory variable (perhaps conditioned on some covariate(s)). In such a situation, the confidence interval on that explanatory variable provides a quick answer: Does the confidence interval include zero? For instance, can we detect in the `Galton` data an effect of the number of children in a household (`nkids`) on the adult `height` of a child? (Speculation on why this might be: Perhaps less food per child is available in large households.) The simplest model is

```{r digits=3}
lm(height ~ nkids, data=Galton) |> conf_interval()
```

The confidence interval on `nkids` does not include zero, so we can fairly say to have detected a connection between `nkids` and `height`. (The effect size is about one-sixth of an inch height loss per additional child in the household.)

TASK 1. Might the link between nkids and height be via sex? Is sex related to `nkids`

```{r}
glm(zero_one(sex, one="F") ~ nkids, data=Galton, family="binomial") |> conf_interval(show_p=TRUE)
```
```{r digits=3}
lm(height ~ nkids + sex + mother + father, data=Galton) |> regression_summary()
```

Focus on cases where
CONFIDENCE INTERVAL JUST BARELY TOUCHES ZERO, p-value is about 0.05





