---
title: 'Lesson 34: Worksheet'
author: "Jane Doe"
---

```{r include=FALSE}
library(math300)
```

In this worksheet, you are going to use the `math300::PIDD` data frame. "PIDD" stands for "Pima Indian Diabetes Database" as you can see by giving the `?PIDD` command.

Your objective is to choose explanatory variables that give a good stratification of the risk of developing diabetes. First, let's turn the `diabetes` categorical variable into a zero-one variable:

```{r}
PIDD <- PIDD |> mutate(developed_diabetes = zero_one(diabetes, one="yes"))
```


The commands building and displaying each model follow:

```{r}
model <- glm(developed_diabetes ~ bmi + insulin, data=PIDD, family="binomial" )
model_plot(model, show_data=FALSE)
```

Change the explanatory variables (use at most three) to get a model specification that stratifies the risk nicely. Once you have a model that you are satisfied with, let's look at it a bit more technically.

```{r}
Scores <- model_eval(model, interval="none") |>
  mutate(diabetes = PIDD$diabetes)
```

The variance of the model output is a nice measure of how much stratification of risk your model gives.

```{r}
Scores |> summarize(vrisk = var(.output))            
```

Choose a threshold for a positive result from your model. This graph will help a lot:

```{r}
ggplot(Scores, aes(x=diabetes, y=.output)) + 
  geom_jitter(alpha=0.3) +
  geom_violin(fill="blue", alpha=0.2, color=NA)
```

Using the threshold you just selected, construct the report on false-positives and false-negatives.

```{r}
threshold <- 0.6 # just demonstrating, but probably not the best choice
Scores |> 
  mutate(test_result = ifelse(.output > threshold, "positive", "negative")) |>
  group_by(diabetes, test_result) |>
  tally()
```

From this report, find the **sensitivity** and **specificity** of your model/threshold. 
