---
title: "Lesson 19: Worksheet"
author: "Jane Doe"
---

```{r include=FALSE}
library(math300)
```

## Objectives

19.1. [Conceptual] Master the use and units of variance and standard deviation in measuring variability.

19.2. [Conceptual] Understand the equivalence between mean and proportion on a zero-one variable.

19.3. [Technical] Use `var()` and `sd()` within `summarize()`

19.4. [Technical] Use `model_plot()` to graph models with one or two explanatory variables.

19.5. [Technical] Use `zero_one()` with `mutate()` to create a zero-one variable.


## Preliminaries: How we will work with R.

In the first half of Math 300Z, the daily student notes were largely structured around "scaffolded" R code, which often involved filling in the blanks. In this second half of 300Z, we will start to use a new way of helping you construct appropriate R command. We call this "*command patterns.* For instance, 

    DF %>% summarize(NM=var(VAR)) 
is a command pattern.

One reason for the shift to the command-pattern style is that there will be only a handful of new patterns in the second half of the course that you'll be using over and over again. Another reason is to help you develop "finger memory" for the most common patterns. An analogy: scaffolding is like GPS navigation which certainly makes it easier to drive but harder to get to know the town. Command patterns are like a paper map, there to help you when you need it.

There is a specific notation for command patterns, which you should memorize. Instead of the blanks used in a scaffold, the command pattern uses a CAPITALIZED abbreviation for the **kind of thing* that should be put in the position. Common kinds of thing are

- `DF`: a data frame, almost always referred to by name.
- `VAR`: a variable in a data frame. Many command patterns involve multiple variables, each of which is referred to by `VAR`. You will replace each `VAR` with the appropriate variable name.
- `VARS`: one or more variable names. When these are the right-hand side of a tilde expression, separate the names with `+` punctuation. When we mean to indicate that there is only one variable, we use `VAR` instead of `VARS`. If we want to say, "use two variables," we would write `VAR + VAR`.
- `MODEL` refers to the name of a model that you have previously constructed with `lm()`.
- `NM` means a name that you will be calling something by. For instance, `NM <- lm(VAR ~ VARS, data=DF)`. Another occasion for using `NM` is as part of an argument to `summarize()` or `mutate()`.
- `[, MORE]` means that you can have multiple additional arguments of the same form as the previous argument. 
- `VALUE` a number, quoted string (e.g., `"red"`), or multiple values inside `c( )`. 
- `MODSPEC` is a model specification, which could equally well be written `VAR ~ VARS`


Anything in a command pattern that is not a CAPITALIZED abbreviation is a specific part of the command to be used as-is. For instance, `lm(VAR ~ VARS, data=DF)` refers explicitly to the `lm()` function whose first argument is a tilde expression and whose second argument is named `data`.

Occasionally, you will refer to a data frame by naming the *package* from which it comes. For example, the `moderndive` package includes (among many others) the `amazon_books` data frame. Think of `amazon_books` as a first name, and `moderndive` as a family name. When you see `PACKAGE::DF` it is meant to indicate, for instance, `moderndive::amazon_books`. (Note that the `::` in the command pattern is to be taken literally; there are two successive colons separating the package name from the name of the data frame.)


## Part 1

Command patterns: 

a. `DF %>% summarize(NM = var(VAR))`  Calculate variance of a variable in a data frame.
b. `DF %>% summarize(NM1 = var(VAR1), NM2 = var(VAR2) [, MORE])
c. `PACKAGE::DF` The name of a data frame within a package.

1. In the `mosaicData::Galton` data frame, find the variance of `mother` and `father`. Give both the numerical value and the units.

::: {.callout-note}
## ANSWER:

```{r}
mosaicData::Galton |> 
  mutate(vmother = var(mother), vfather = var(father))
```

The units for both are `inches-squared" since the variables themselves have units "inches."
:::


2. In the `moderndive::amazon_books` data frame, find the variance of `list_price` and `num_pages`. Give both the numerical value and the units.

::: {.callout-note}
## ANSWER:

```{r}
moderndive::amazon_books |>
  mutate(vprice = var(list_price), vpages = var(num_pages))
```

The units of `list_price` are dollars, so the variance has units "square-dollars".

`num_pages` is dimensionless, so the variance is also dimensionless. 
:::


3. Calculate the variance of `sex` from `Galton`. If something goes wrong, explain why.

::: {.callout-note}
## ANSWER: 

```{r error=TRUE}
Galton |> mutate(vsex = var(sex))
```

`sex` is a categorical variable. There's no such thing as the variance of a categorical variable.
:::

## Part 2

Command patterns:

a. `NM <- lm(VAR ~ VARS, data = DF)`
b. `lm(VAR ~ VARS, data=DF) %>% conf_interval()` 
c. `lm(MODSPEC, data=DF) %>% conf_interval()` means the same as (b).


1. (Easy, no computing needed.) What kind of a thing is `conf_interval()`. (Hint: It's the same kind of thing as `lm()`.)

::: {.callout-note}
## ANSWER:

`conf_interval()` is a *function*.
:::


2. Using the `moderndive::amazonbooks` data frame, fit the model `list_price ~ num_pages`: 
    i. What are the units of the "(Intercept)" coefficient?
    i. Report the coefficient on `num_pages`. Give both the numerical bounds and the units.
    

::: {.callout-note}
## ANSWER:

The intercept always has the same units as the response variable. Here, that's dollars.

```{r}
lm(list_price ~ num_pages, data = moderndive::amazon_books) |>
  conf_interval()
```

Coefficient on `num_pages`: .02 dollars per page. Multiplying the coefficient by the number of pages will give dollars: the units of the response variable.
:::


3. Similar to (2) but with the model `list_price ~ numpages + hard_paper`
    i. What does the term `hard_paperH` refer to? 
    i. According to the coefficients, is a hardcover book any more expensive (on average) than a softcover book? 

::: {.callout-note}
## ANSWER:

```{r}
lm(list_price ~ num_pages + hard_paper, data = moderndive::amazon_books) |>
  conf_interval()
```

`hard_paperH` refers to the `H` level of the `hard_paper` variable. According to the model, a hardback costs $4.96 more than a paperback, on average.
:::

4. Store the model you created in (3) under the name `mod3`. We'll use it in the next part. For your answer, put the R command you used to store the model as `mod3`.


::: {.callout-note}
## ANSWER:

Note that we are asked to store the model itself, not the confidence interval.

```{r}
mod3 <- lm(list_price ~ num_pages + hard_paper, data = moderndive::amazon_books) 
```

:::

    
## Graphics review

Command patterns:

a. `ggplot(DF, aes(x=VAR, y=VAR)) + geom_jitter()`
b. `ggplot(DF, aes(x=VAR, y=VAR)) + geom_jitter() + geom_violin(fill="blue", alpha=0.3)`
c. `ggplot(DF, aes(x="all", y=VAR)) + geom_jitter()`
c. `model_plot(MODEL, x=VAR)`
d. `model_plot(MODAL, x=VAR, color=VAR)`

1. Make a jitter plot of `list_price ~ hard_paper` from `moderndive::amazon_books`.

::: {.callout-note}
## ANSWER:

```{r}
moderndive::amazon_books |>
  ggplot(aes(x=hard_paper, y = list_price)) + 
  geom_jitter(alpha=0.5)
```
:::

2. Using your command from (1), add a new layer: `+ geom_violin(fill="blue", alpha=0.3)`

::: {.callout-note}
## ANSWER:

```{r}
moderndive::amazon_books |>
  ggplot(aes(x=hard_paper, y = list_price)) + 
  geom_jitter(alpha=0.5) +
  geom_violin(fill="blue", alpha=0.3)
```

:::

3. Use `model_plot()` to draw a picture of `mod3`. Set `x=hard_paper` and `num_pages=200`. What do you think the horizontal line segments refer to?

::: {.callout-note}
## ANSWER:

```{r warning=FALSE}
model_plot(mod3, x=hard_paper, num_pages=200)
```

The vertical position of the horizontal lines indicates the model output for books with 200 pages.

:::

4. Repeat (3), but remove `num_pages = 200`. Instead, set `x=num_pages` and `color=hard_paper`. Explain the meaning of the line segments in everyday terms. 

::: {.callout-note}
## ANSWER:

```{r warning=FALSE}
model_plot(mod3, x=hard_paper, color=num_pages)
```

The parallel line segments in each column show the model output for books with 200, 400, and 600 pages respectively.

:::
