---
title: "Math 300R NTI Lesson `r (lesson <- 20)`"
subtitle: "title goes here"
author: "Prof. Danny Kaplan"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output:
  html_document:
    theme: lumen
    toc: yes
    toc_float: yes
    css: NTI.css
  pdf_document:
    toc: yes
---


```{r setup, include=FALSE}
library(knitr)
library(kableExtra)
library(ggformula) # Change plots to ggplot2
library(tibble)
knitr::opts_chunk$set(echo = TRUE)
```

## Objectives

#. Understand that gaming is a way of improving our skills and identifying potential opportunities and problems. 

#. Characterize the magnitude of random noise.

#. Distinguish between a sample, a summary of a sample, and a sample of summaries of samples.

## Reading

TBD




## Lesson

::: {.callout-note icon=false}
## Setup

The software for generating the simulations, and which contains a library of different scenarios, is in the package `{dagsim}`. The `{mosaic}` package contains functions for calculating statistics and **iteration**. 

```{r message=FALSE, warning=FALSE}
library(dagsim)
library(mosaic)
```

::: {.callout-tip icon=false}
## Example
1. Roll a die 1000 times. How often does each possible result appear?
```{r}
Sim <- dagsim(list(x ~ each(sample(1:6))), nrow=1000)
Sim  %>% group_by(x) %>% summarize(count = n())
```


1. Generate 1000 rolls of the sum of two dice. How often does each possible result appear?

```{r}
Sim <- dagsim(list(x ~ each(sum(sample(1:6, 2, replace=TRUE)))), nrow=100)
Sim %>%  
  dplyr::group_by(x) %>%
  summarize(frac = n()/nrow(.)) %>% 
  print.data.frame()
```

Question: What do your results tell you about the possible results when summing 2 dice? 

2. Do the same but for summing 6 dice. 

Question: What do your results tell you about the possible results when summing 6 dice? Do the data show you that it is possible to get a result of 36?  

This illustrates something about "inference": the data do not always point you directly to the truth.

Question: How would you modify the simulation to get a better handle on the truth. 
:::

::: {.callout-note icon=false}
## Measuring noise
Create a DAG where the X variable is simply noise: `eps()`

```{r}
Sim <- dagsim(list(x ~ eps()), nrow=100)
Sim %>% summarize(sd(x), mean(x))
```

- The argument to `eps()` sets the size of the noise. Confirm that this claim is true.
:::

Sometimes we will want to repeat a simulation over and over again, collecting the results. Use `do()` for this:

```{r}
# one trial
dagsim(list(x ~ eps()), nrow=100) %>% 
  summarize(sd(x), mean(x))

# five trials
do(5) *
  {dagsim(list(x ~ eps()), nrow=100) %>% 
  summarize(s = sd(x), m = mean(x))}
```

- Not all the rows are the same. Why not?
- None of the standard deviations are exactly 1. Does this mean the simulation is not working? 

A typical problem in statistical inference is to determine, from just one trial, whether the results are consistent or not with a proposed mechanism.

## Learning Challenges

:::: {.callout-note icon=false}
## LC 20.1 (Objective 2)

This is just a template.

::: {.callout-tip}
## Solution


The solution will go here.
:::
::::

## Documenting software

  * File creation date: `r Sys.Date()`
  * `r R.version.string`
  * `tidyverse` package version: `r packageVersion("tidyverse")`

