---
title: "Math 300R NTI Lesson `r (lesson <- 24)`"
subtitle: "Effect size"
author: "Prof. Danny Kaplan"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output:
  html_document:
    theme: lumen
    toc: yes
    toc_float: yes
    css: NTI.css
  pdf_document:
    toc: yes
---


```{r setup, include=FALSE}
library(knitr)
library(kableExtra)
library(ggformula) # Change plots to ggplot2
library(tibble)
knitr::opts_chunk$set(echo = TRUE)
```

## Objectives

#. Estimate an effect size from a regression model of one and two variables.

#. Construct a confidence interval on the effect size.

#. Gaming: Evaluate whether confidence interval indicates that estimated effect size is consistent with simulation.


## Reading

TBD


## Lesson

:::: {.callout-note icon=false}
## In-class example

```{r}
3 + 2
```
::::


:::: {.callout-important icon=false}
## In-class activity

```{r}
date()
```
::::

This is just a template.

## Learning Challenges

:::: {.callout-note icon=false}
## LC 19.1 (Objective 2)

This is just a template.

::: {.callout-tip}
## Solution


The solution will go here.
:::
::::

## Documenting software

  * File creation date: `r Sys.Date()`
  * `r R.version.string`
  * `tidyverse` package version: `r packageVersion("tidyverse")`

