## Math 300Z In-class activity

```{r include=FALSE}
library(math300)
```

## Intervals by eye

::: {.content-hidden when-format="pdf"}

In this activity, you are given some point plots of data: y vs x. Your job is to

1. sketch in an appropriate model fitted (by eye) to the data.
2. add a prediction band showing for each value of x what is the **prediction** interval
3. transform the prediction band into a **confidence** band.

TIPS:

a. The fitted model will be a line or curve, or in the case of Model 3, two lines.
b. The bounds of the prediction band will be more-or-less parallel to the fitted model, but should include roughly 95% of the $n$ points in the plot.
c. The confidence band is narrower than the prediction band by a factor of $1/\sqrt{n}$.

:::



::: columns

::: {.column width="50%"}

### Model 1: A straight-line model `y ~ x`

```{r echo=FALSE, fig.height=3, fig.width=6}
set.seed(102)
One <- dag_make(
  x ~ unif(),
  y ~ 4 - 2.5*x + exo()
)
Sone <- sample(One, size=100)
gf_point(y ~ x, data=Sone) %>%
  gf_labs(subtitle="n=100") %>%
  gf_lims(y=c(-2,7))
```
:::


::: {.column width="50%:}
### Model 2: A sine-wave model, `y ~ sin(x)`

```{r echo=FALSE, warning=FALSE, fig.height=3, fig.width=6}
set.seed(102)
Two <- dag_make(
  x ~ unif(0, 10),
  y ~ 3 + 4* sin(x) + exo()
)
Stwo <- sample(Two, size=400)
gf_point(y ~ x, data=Stwo) %>%
  gf_labs(subtitle="n=400") %>%
  gf_lims(y=c(-2,7))
```
:::

:::



### Model 3: A function of two variables, `y ~ x + group`

```{r echo=FALSE, fig.height=3, fig.width=6}
set.seed(102)
Three <- dag_make(
  x ~ unif(0, 10),
  group ~ binom(0, labels=c("group A", "group B")),
  y ~ 1 + 5*(group=="group A") + 2 * x + exo(1.5)
)
Sthree <- sample(Three, size=200)
gf_point(y ~ x, data=Sthree, shape=~group) %>%
  gf_labs(subtitle="n=200") 
```

