---
title: "Math 300R Lesson `r (lesson <- 30)` Reading Notes"
subtitle: "Confounding"
author: "Prof. Danny Kaplan"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output:
  html_document:
    theme: lumen
    toc: yes
    toc_float: yes
    css: NTI.css
  pdf_document:
    toc: yes
---

```{r setup, include=FALSE}
source("../_startup.R")
```

::: {.callout-note icon=false}
## Causal Caution

Just because you've calculated an effect size doesn't mean that you have captured any sort of causal relationship between the variables. To illustrate, use `dag01` and fit two different models: `y ~ x` and `x ~ y`.

```{r}
Sample <- sample(dag01, size=500)
lm(y ~ x, data = Sample)
lm(x ~ y, data = Sample)
```

You can't tell from these coefficients whether `x` causes `y` or vice versa (or something entirely different). The words "correlation" or "association" are used when we don't want to claim that there is a causal connection. Many statisticians will only use those words unless the data come from an **experiment**. 

We're going to use causal language ("relationship", "effect," etc.) because that is often the matter of concern to decision making. But using language doesn't doesn't make the connection causal.

:::
