---
title: "Instructor Teaching Notes for Lesson 25"
subtitle: "Math300Z"
author: "Daniel Kaplan"
date: "Revised 03/08/2023"
---


```{r include=FALSE}
library(math300)
```

## Review of Lesson 24

In the last lesson, we considered **effect size**, a way to summarize a model to indicate the strength and direction of the influence of an explanatory variable with respect to a response variable. (As always, sample statistics such as the effect size are **according to the model**. A different model may give you different results, so the choice of model specification is important and should be carefully considered.)

Here's a possible DAG for the height of a child.

```{r}
height_dag <- dag_make(
  mother_genetics ~ exo(),
  father_genetics ~ exo(),
  exo_dad ~ exo(),
  exo_mom ~ exo(),
  mother_height ~ mother_genetics + exo_mom + exo(),
  father_height ~ father_genetics + exo_dad + exo(),
  health ~ exo(),
  nutrition ~ exo(),
  child_sex ~ exo(),
  child_height ~ mother_genetics + father_genetics + child_sex + nutrition + health + exo()
)
set.seed(106); dag_draw(height_dag, vertex.size = 13,
                        vertex.label.cex=0.75, 
                        edge.arrow.size=.25 )
```

We talk casually about the *effect size of child's height with respect to mother's height*, but there is no causal flow between mother's height and child's height. What we mean is the effect of a change in the mother's genetics that leads to a 1 inch change in mother's height and a corresponding change in child's height.

The concept of an *effect size* is that all the other nodes are held constant when we change the mother's genetics, including `health`, `nutrition`, `sex`, and father's genetics. 

There's no way to do this; there aren't two identical mothers to compare who differ only in their height-related genetics and can produce children who have exactly the same father's genetics. 

Instead, what we think about is comparing two children from two similar mothers (differing only in height genetics) with similar father genetic contribution, nutrition, health, ....

The model specification `height ~ mother + father + sex` does this comparison of similars in a mathematical way. But the model output is not the only thing that changes when we generate this mathematical child's height from two different height mothers.

The **residuals** from the model indicate the magnitude of all the other influences that we haven't been able to hold constant. 

```{r}
height_mod <- lm(height ~ mother + father + sex, data=Galton)
Values <- model_eval(height_mod)
ggplot(Values, aes(x=mother, y=.resid)) + geom_jitter(alpha=0.5)
```

1. Is there any clear pattern to the residuals? 

2. If we were making a prediction of a daughter's height, should we just look at the model value based on the parents' heights, or should we take into account the residuals.

## Proper form for a prediction

Assign a probability to every possible outcome. 

```{r}
ggplot(Values, aes(x=" ", y=.resid)) + geom_violin(alpha=0.4, fill="blue") +
  geom_jitter(alpha=0.2, width=.15) + xlab("Residual") + ylab("") +
  geom_hline(yintercept=c(-4.13, 4.04), color="red", alpha=0.5)
Values |> summarize(m = mean(.resid), sd=sd(.resid), 
                    q2.5=quantile(.resid, 0.025),
                    q97.5 = quantile(.resid, 0.975))
```

## Realistic predictioç

Precipitation runoff: It rains. Some of the water is absorbed, some runs off and ends up in a nearby river. The runoff is measured as the depth of water over the entire catchment basin that ends up in the river. Give a forecast of a storm, we might want to know if a flood is likely.

Data for the Monocacy River in Maryland, close to where I grew up.

RESULT FOR THE BIGGEST RAINFALL DEPENDS ON how logarithms are used.

```{r}
ggplot(Monocacy_river |> filter(precip > 2), aes(x=log(precip), y=log(runoff))) +
  geom_point() +
  geom_lm(interval="prediction") +
  geom_lm(interval="confidence", fill="blue") # +
 # geom_abline(slope=1.08, intercept=-1.07)
```
