---
title: "Math 300 Lesson 26 Notes"
subtitle: "Mechanics of prediction"
author: "YOUR NAME HERE"
date: "`r format(Sys.time(), '%B, %Y')`"
output:
  pdf_document:
    toc: yes
  html_document:
    theme: lumen
    toc: yes
    toc_float: yes
---

```{r setup26, include=FALSE}
library(math300)
library(mosaic)
library(knitr)
library(kableExtra)
library(math300)
# math300::setup_student_notes()
```

## Overview

It is often helpful to think of a model as a function that translates inputs into an output. An "**effect size**" is one way of summarizing that model. It is merely the change in output that results from a 1-unit change of input of a specified explanatory variable.

### Reading

[Lesson **26** from *LST*](https://dtkaplan.github.io/Math-300R/Textbook/Reading-notes-lesson-26.html)

### Objectives



[Updated list of objectives](../Objectives/Obj-lesson-26.html)

### Libraries

```{r warning=FALSE,message=FALSE}
library(mosaic)
library(math300)
data("Galton", package="mosaicData")
data("Home_utilities", package="mosaicCalc")
data("Body_fat", package="mosaicCalc")
```

----------

## Lesson

### Exercise 26.2

University-level students are well familiar with two kinds of tests: tests used to assign a grade in a course and so-called "standardized tests" used to regulate college and graduate-school admissions. Famous examples are the SAT and the ACT tests used by many college admissions offices.

There is controversy about the use of standardized tests to guide college admission. Much of this controversy stems from disappointment about the lack of diversity at many elite colleges. This is a complex social and political issue.

Our concern in this exercise is with another question that is fundamentally statistical: Do the SAT and ACT have *predictive value*?

The College Board, the organization that owns and administers the SAT, published a [report in 2019](https://satsuite.collegeboard.org/media/pdf/national-sat-validity-study-overview-admissions-enrollment-leaders.pdf) emphasizing this claim.

> *The SAT is strongly predictive of college success; students with higher SAT scores are more likely to have higher grades in college.*

In this exercise, we are going to use DAG-simulated data that has the same statistical properties as the SAT/college-grade data that formed the basis for the College Board's conclusion. 

The `dag_satgpa` DAG defined the `{math300}` package generates two columns of output, a (simulated) SAT score between 400 and 1600 points, and a (simulated) GPA between 0 and 4.0. 


#### Task 2.1

Generate a sample of size $n=1000$ from `dag_satgpa`. Use that sample to train the model `gpa ~ sat`.

```{r}
# For Task 2.1
 Scores <- sample(dag_satgpa, size=1000)
 Model <- lm(gpa ~ sat, data = Scores)
```

Graph the data and the model using `model_plot()`:

```{r}
model_plot(Model, x = sat)
```

Explain how the plot is consistent or inconsistent with the claim that SAT is "strongly predictive" of GPA. 

*Your answer*:

Add a confidence band to the plot of the model. (Use an additional argument to `model_plot()`, namely, `interval="confidence"`.) Does this substantially change your previous conclusion about the predictive ability of SAT?

*Your answer*:

#### Task 2.2

Using the graph with the confidence interval from `model_plot()`, say what you think should be the predicted GPA for a student who scores 1200 on the SAT? Explain what in the graph leads you to this conclusion. 

*Your initial answer*:

Later on, you may come to reconsider your answer. But we want to record your initial answer here, so that later you can see whether your understanding of prediction has changed.

#### Task 2.3

A proper prediction should always come with a statement about the uncertainty. Since we have been working with *confidence intervals* over the last few Lessons, you may be inclined to think that a confidence interval is a way to measure the uncertainty in a prediction. But that is **absolutely wrong**!

```{r}
model_plot(Model, x = sat, interval="prediction")
```



#### Task 2.4

```{r}
Scores <- sample(dag_satgpa, size=10000) %>%
  mutate(above = ifelse(gpa > 3.0, "above", "below"))
ggplot(Scores, aes(x=sat, y=above)) + 
  geom_jitter(alpha=0.1)
mod <- glm(zero_one(above, one="above") ~ sat,
           data=Scores, family="binomial")
model_plot(mod, x = sat)
```
